---
title: "ADaM Specification"
vignette: >
  %\VignetteIndexEntry{adam_specification}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Overview

The ADaM specification required by mighty follows the mighty.metadata format. This article discusses the structure and fields used to define ADaM datasets. Additional fields beyond those discussed here are ignored by mighty.

## General structure

Each ADaM domain is defined in its own YAML file with these main sections:

1.  **Domain identifier** - The ADaM domain name
2.  **Primary keys** - Unique identifiers for the dataset
3.  **Label** - Label of the ADaM dataset
4.  **Class** - CDISC class of the dataset
5.  **Structure** - Text description of the structure of the dataset
3.  **Population** - How to build the base table from SDTM domains
4.  **Columns** - Which columns to include and how to create them
5.  **Rows** - Operations that add, modify, or delete rows
6.  **Parameters** (BDS only) - Parameter-level derivations

# Domain identifier

The domain identifier is specified at the top level of the YAML file.

## Required Field

| Field | Description            | Example |
|-------|------------------------|---------|
| `id`  | Name of the ADaM domain | `ADLB`  |

## Example

``` yml
id: ADLB
```

# Primary keys

Define the primary key variables that uniquely identify each row in your ADaM dataset.

## Required Field

| Field  | Description                                    | Example              |
|--------|------------------------------------------------|----------------------|
| `keys` | Array of variable names forming the primary key | `[USUBJID, PARAMCD]` |

## Example
``` yml
keys: [USUBJID, PARAMCD, AVISIT]
```

# Label

A descriptive label for the ADaM domain.

## Required Field

| Field   | Description                        | Example                        |
|---------|------------------------------------|--------------------------------|
| `label` | Human-readable domain description | `Laboratory Analysis Dataset` |

## Example
``` yml
label: Laboratory Analysis Dataset
```

# Class

The ADaM dataset class according to ADaM standards.

## Required Field

| Field   | Description                    | Example                   |
|---------|--------------------------------|---------------------------|
| `class` | ADaM class of the dataset      | `BASIC DATA STRUCTURE`    |

## Common values

- `SUBJECT LEVEL ANALYSIS DATASET` - For ADSL
- `BASIC DATA STRUCTURE` - For BDS datasets (ADLB, ADVS, etc.)
- `OCCURRENCE DATA STRUCTURE` - For OCCDS datasets (ADAE, ADCM, etc.)

## Example
``` yml
class: BASIC DATA STRUCTURE
```

# Structure

A description of the dataset structure that specifies the grain/level of the data.

## Required Field

| Field       | Description                          | Example                                                   |
|-------------|--------------------------------------|-----------------------------------------------------------|
| `structure` | Description of dataset granularity   | `One record per subject per parameter per analysis visit` |

## Example
``` yml
structure: One record per subject per parameter per analysis visit
```

# Population

This section defines how to create the base population for your ADaM table from source SDTM domains and apply filters.

## Structure

The `population` section has two subsections:

| Field    | Description                                        |
|----------|----------------------------------------------------|
| `base`   | Array of base SDTM domains with domain-specific filters |
| `global` | Array of filters applied to the combined dataset   |

## Filter execution order

Understanding when filters are applied is essential for writing correct specifications. The following describes how filtering is implemented in the generated programs. The `SRC_` column mentioned in step 2 is an internal implementation detail added and removed automatically by mighty — it never appears in the specification or the final dataset:

1. **Read data**: Base domains are read from SDTM
2. **Add source tracking**: A temporary `SRC_` column is added to each domain to track its origin (only when domain-specific filters are present)
4. **Apply domain-specific filters**: Filter expressions from `base.filter` are applied using the `SRC_` column to determine which conditions apply to which source domain
5. **Drop `SRC_` column**: The temporary source tracking column is removed
6. **Join external data**: If global filters reference external domains (e.g., `ADSL`), those datasets are joined
7. **Apply global filters**: Filter expressions from `global.filter` are applied sequentially to the combined dataset
8. **Continue with derivations**: Column and row operations proceed

Domain-specific filters allow different inclusion criteria for different source domains, while global filters apply uniform criteria across the combined dataset.

## Base domains

Each entry in the `base` array specifies a source domain:

| Field    | Description                                     | Notes |
|----------|-------------------------------------------------|---------------|
| `domain` | SDTM domain name                                                    | Case-sensitive |
| `depends`| Columns needed for filtering this domain                            | Can be empty array `[]` |
| `filter` | Filter specification. Must be an R expression that can be evaluated | Can be `NA` |

### Domain-specific filtering

Domain-specific filters allow you to apply different inclusion criteria to different source domains before they are combined. Each domain's filter expressions are combined with `&` (AND logic) and wrapped with `(SRC_ == 'domain_name' & conditions)`. Multiple domains' filter conditions are then combined with `|` (OR logic).

This is useful when:

- Different domains need different inclusion criteria
- You want to exclude specific records from one domain but keep all records from another
- Filtering logic depends on the source domain

## Global filters

Global filters are applied after row-binding and domain-specific filtering.

| Field    | Description                         | Notes                                |
|----------|-------------------------------------|--------------------------------------|
| `depends`| Array of columns needed for global filtering | Include domain prefix (e.g., `ADSL.SAFFL`) when referencing a column outside the current domain |
| `filter` | Filter specification. Must be an R expression that can be evaluated        | Applied to combined dataset. |

### Using external domains in global filters

Global filters can reference columns from external ADaM or SDTM domains by:

1. Listing required columns in `global.depends` with domain prefix (e.g., `ADSL.SAFFL`)
2. Writing filter expressions that reference the columns without the prefix (e.g., `SAFFL == "Y"`)

mighty automatically handles the join operation using the primary keys defined in `_mighty.yml` `external_data`.

## Simple example

``` yml
population:
  base:
    - domain: LB
      filter: NA
      depends: NA
  global:
    - filter: NA
      depends:
        - NA
```

## Complex example

``` yml
population:
  base:
    - domain: LB
      filter: NA
      depends: NA
    - domain: XL
      filter:
        - 'grepl("_CONV", TOPICCD)'
      depends:
        - TOPICCD
  global:
    - filter: 'LBCAT != "ANTIBODIES"'
      depends:
        - LBCAT
    - filter: '!ARMNRS %in% c("SCREEN FAILURE")'
      depends:
        - ADSL.ARMNRS
```

## Key Points

-   Multiple base domains are row-bound together
-   Base filters apply to specific domains before combining
-   Global filters apply to the combined dataset
-   Column dependencies must include domain prefixes when from external domains (e.g., `ADSL.ARMNRS`)

# Columns

Define which columns to include in your ADaM dataset and how to create them.

Column operations cannot modify the number of rows in the ADaM table. To modify rows, use the `rows` section (see below).

The `columns` section is an array where each element defines one or more columns.

## Fields

Each column specification uses different fields depending on how the column is created:

| Field       | Description                                                   | Notes |
|-------------|---------------------------------------------------------------|-------|
| `id`        | Name of the column in the final dataset                       | Required |
| `method`    | Source location for simple operations                         | Use for copying/renaming columns |
| `component` | Component definition for derivations                          | Object with `id` and optional `with` |
| `depends`   | IDs of row or parameter operations that must complete first   | Array of strings (e.g., `rows.BASELINE_FLAG`, `parameters.BMI`) |

### Component object

When using `component`, it has these subfields:

| Field | Description                                      | Notes |
|-------|--------------------------------------------------|-------|
| `id`  | Name of standard component or path to custom one | Required |
| `with`| Parameters to pass to the component              | Object of key-value pairs |

## Examples

### Simple column operations

1.  Adding a column from the base domain:

    ``` yml
    columns:
      - id: USUBJID
    ```

    Retains the USUBJID variable from your base SDTM domain (e.g., LB.USUBJID).

2.  Adding a column from another ADaM domain

    ``` yml
    columns:
      - id: SAFFL
        method: ADSL.SAFFL
    ```

    Joins the SAFFL variable from ADSL onto your current ADaM table using common keys.

3.  Adding a column from another ADaM domain and renaming it

    ``` yml
    columns:
      - id: ARM
        method: ADSL.PLANNED_ARM
    ```

    Merges PLANNED_ARM from ADSL and names it as ARM in your table. The join is done via the common keys of the current dataset and ADSL.

4.  Copying a column from the base domain with a new name

    ``` yml
    columns:
      - id: LBSTRESN
      - id: AVAL
        method: LBSTRESN
    ```

    Creates a new column AVAL by copying the content of LBSTRESN from the base domain. Note that LBSTRESN must also be specified in the columns section. Both columns will exist in the resulting dataset.

5.  Renaming a column from the base domain

    ``` yml
    columns:
      - id: AVAL
        method: LBSTRESN
    ```

    Renames `LBSTRESN` from the base domain to `AVAL`. Unlike copying (example 4), the source column `LBSTRESN` is **not** listed separately in `columns`, so only the renamed column `AVAL` appears in the final dataset.

### Derived columns

Derived columns use code components to perform data transformations. A code component can be:

1.  Standard code component (Mustache template)
2.  Custom component (plain R script)
3.  Custom component (Mustache template)

#### Examples (continued):

6.  Adding a derived column using a standard code component

    ``` yml
    columns:
      - id: ARM_GRP1
        component:
          id: arm_group_01
    ```

    Creates `ARM_GRP1` using the standard code component `arm_group_01`. The component's metadata defines which columns it depends on.

7.  Adding a derived column using a custom code component (plain R script)

    ``` yml
    columns:
      - id: ARM_GRP1
        component:
          id: path/to/code_component/arm_group_01.R
    ```

    For custom code components, provide the file path. Custom R scripts should contain the data manipulation code directly as executable script code, and metadata to declare dependencies and columns created by the component. For details, see `vignette("code_components")`.

8.  Adding a derived column using a custom code component (Mustache template)

    ``` yml
    columns:
      - id: ARM_GRP1
        component:
          id: path/to/code_component/arm_group_01.mustache
    ```

    Custom Mustache templates allow parameterization (see example 10).

9.  Adding multiple derived columns with a single standard code component

    ``` yml
    columns:
      - id: NEWFL01
        component:
          id: newfl_01
      - id: NEWREA01
        component:
          id: newfl_01
    ```

    Adds both `NEWFL01` and `NEWREA01` columns by calling the standard code component `newfl_01`. The component must be designed to output both variables.

10. Adding a derived column using a parameterized code component

    ``` yml
    columns:
      - id: AGE_GRP1
        component:
          id: age_group_01
          with:
            cut_points: c(20, 45)
    ```

    Creates `AGE_GRP1` using `age_group_01` code component with custom parameters. Parameters can only be passed to Mustache templates (standard components and custom Mustache components), not to plain R script components.

11. Adding multiple derived columns with a single, parameterized code component

    ``` yml
    columns:
      - id: A
        component:
          id: fn_AB
          with:
            param_1: 100
            param_2: "This is a regular string"
            param_3: min(1, 100)
      - id: B
        component:
          id: fn_AB
          with:
            param_1: 100
            param_2: "This is a regular string"
            param_3: min(1, 100)
    ```

    Adds columns `A` and `B` using the same parameterized code component `fn_AB`. When one code component creates multiple columns, all columns **must** use identical parameter values.

12. Adding a derived column that has a row dependency

    ``` yml
    columns:
      - id: BASETYPE
        component:
          id: basetype_derivation
        depends:
          - rows.BASELINE_FLAG
    ```

    Creates `BASETYPE` using `basetype_derivation` code component, but only after the row operation `BASELINE_FLAG` completes. Note the `rows.` prefix for row dependencies.

## Summary of column rules

1.  A single component can produce multiple columns in the ADaM dataset. All columns generated by that component must have a corresponding entry in the columns section. The number of columns a component produces is determined by the component's code, not by the ADaM specification.
2.  All parameters expected by the component need to be specified in the `with` section of the component specification.
3.  Dependencies must use appropriate prefixes: `rows.` for row operations (e.g., `rows.BASELINE_IMPUTATION`) or `parameters.` for parameter-level operations in BDS datasets (e.g., `parameters.BMI`).

## Component reuse rules

Components can appear more than once in a domain's `columns` section. There are two distinct scenarios.

### One component, multiple output columns

A component with multiple `@outputs` appears under several column entries in the YAML. Because mighty calls the component once, all entries must use **identical** `with:` values. See examples 9 and 11 above for this pattern.

### Same component, different parameters

The same template can be invoked multiple times with different parameter values to produce different output columns. Mighty calls the component once per unique parameter set. This is valid as long as each invocation produces disjoint output columns — **no two invocations may share an output column**.

### Overlapping outputs cause an error

When two invocations with different parameters produce overlapping output columns, the derivation is ambiguous and mighty raises an error.

Consider a component that derives both a date and its imputation flag — it produces two output columns from one invocation:

```{r}
#| echo: false
#| output: false
library(mighty)
component_dir <- tempfile()
dir.create(component_dir)
component_file <- file.path(component_dir, "dtc_to_dt.mustache")
writeLines(
  "
#' @title Convert DTC to DT
#' @description Derives a date and date flag from a DTC variable
#' @param dtc_var Name of the source DTC variable
#' @param dt_var Name of the output date variable
#' @param dtf_var Name of the output date flag variable
#' @type derivation
#' @depends ADLB USUBJID
#' @outputs {{{dt_var}}}
#' @outputs {{{dtf_var}}}
#' @code
ADLB <- ADLB |>
  dplyr::mutate(
    {{{dt_var}}} = as.Date({{{dtc_var}}}),
    {{{dtf_var}}} = ifelse(is.na({{{dtc_var}}}), 'Y', NA_character_)
  )
",
  con = component_file
)

invalid_columns_yml <- glue::glue(
  "
columns:
  - id: USUBJID
  - id: ASTDT
    component:
      id: {component_file}
      with:
        dtc_var: LBDTC
        dt_var: ASTDT
        dtf_var: ASTDTF
  - id: ASTDTF
    component:
      id: {component_file}
      with:
        dtc_var: LBENDTC
        dt_var: ASTDT
        dtf_var: ASTDTF
"
)
```

If the same component is used twice with different parameters but both uses produce the same output columns, mighty cannot tell which parameter values should be used:

```{r}
#| echo: false
#| output: asis
columns_display <- gsub(
  component_file,
  "dtc_to_dt.mustache",
  invalid_columns_yml
)
cat("```yml", trimws(columns_display), "```", sep = "\n")
```

Mighty will abort with an error:

```{r}
#| error: true
#| echo: false
yml <- paste0(
  "
id: ADLB
label: Laboratory Analysis Dataset
class: BASIC DATA STRUCTURE
structure: One record per subject per parameter per analysis visit
keys: [USUBJID]
population:
  base:
    - domain: LB
      depends:
        - NA
      filter: NA
  global:
    - filter: NA
      depends:
        - NA
",
  invalid_columns_yml
)

study_dir <- tempfile()
dir.create(study_dir)
writeLines(yml, file.path(study_dir, "adlb.yml"))
writeLines("external_data:\n  - id: DM\n    keys: [STUDYID, USUBJID]", file.path(study_dir, "_mighty.yml"))
writeLines("study_id: demo_study", file.path(study_dir, "_study.yml"))

generate_adam_code(
  adam_specifications = study_dir,
  path_connector_config = file.path(tempdir(), "_connector.yml"),
  check_cross_domain_adam_dependencies = FALSE
)
```

The fix is to ensure that each invocation of the component (that is, with different parameters) produces a distinct set of output columns.

### Cross-domain reuse

Cross-domain reuse is not affected by these rules. The same component can appear in different domains with different parameters even if the output column names overlap, because each domain is independent.

## No intermediate columns

The YAML specification describes only columns that appear in the final ADaM dataset and in the define.xml. There is no mechanism to declare intermediate or temporary variables — columns used during derivation but excluded from the final output. This affects two situations:

**Source columns not in the final dataset.** A component may need a column from a source domain that does not belong in the output (e.g., `LBDTC` for deriving `ADT`). Declare these as `@depends` in the component's metadata header. Mighty ensures the column is read and available at execution time without it appearing in the specification. See [Handling external dependencies](code_components.html#handling-external-dependencies) for details.

**Intermediate derived values shared across final columns.** In traditional ADaM programming, a temporary column might be calculated once, used by several downstream derivations, and then dropped. Mighty does not support this pattern — there is no way to declare a column that exists only during execution. Instead, either group the final columns that share the intermediate calculation into a single component (using multiple `@outputs`), or recalculate the intermediate value within each component that needs it.
Recalculating intermediate values across multiple components may reduce computational efficiency, but this is not a primary concern. The paramount goal of ADaM programs is clarity and transparency for regulatory review. Therefore, group derivations into shared components only when it improves the logical flow and understanding of the analysis, not simply to optimize performance.
# Row operations

Row operations are specified in the `rows` section of the ADaM domain YAML as an array.

Like column derivations, row operations can use either standard components, custom Mustache templates, or custom R components.

Row operations have an `id` field (user-specified identifier) and a `component` field. The `id` is needed because the same component can be used with different parameter values, and the `id` allows you to refer to a specific implementation when specifying dependencies.

## Examples

1.  Simple row operation

    ``` yml
    rows:
      - id: BASELINE_FLAG
        component:
          id: add_baseline_visit
    ```

    Executes the `add_baseline_visit` standard code component. The specific actions depend on the code component's implementation.

2.  Row operation with parameterized code component

    ``` yml
    rows:
      - id: BASELINE_FLAG
        component:
          id: add_baseline_visit
          with:
            first_treatment_visit: n
    ```

    Specifies `add_baseline_visit` and passes the parameter `first_treatment_visit = n`.

3.  Multiple row operations referencing the same code component

    ``` yml
    rows:
      - id: LOCF_IMPUTATION
        component:
          id: impute_values
          with:
            method: LOCF
      - id: NOCB_IMPUTATION
        component:
          id: impute_values
          with:
            method: NOCB
    ```

    Uses the same code component `impute_values` twice with different parameter values. Unlike column operations, row operations allow the same code component to be called multiple times with different parameters.

4.  Row operation depending on another row operation

    ``` yml
    rows:
      - id: BASELINE_FLAG
        component:
          id: add_baseline_visit
      - id: CHANGE_FROM_BASELINE
        component:
          id: calculate_chg
        depends:
          - rows.BASELINE_FLAG
    ```

    Runs `calculate_chg` only after `BASELINE_FLAG` is executed. Use this when operations must happen in a specific order. Note the `rows.` prefix in the dependency.

# Parameters

The `parameters` section is used primarily in BDS (Basic Data Structure) datasets to define parameter-level operations. These operations create new parameter records (e.g., adding BMI as a new PARAMCD) by deriving values across existing parameters.

Parameters are specified as an array, similar to rows and columns.

## Examples

1.  Simple parameter operation

    ``` yml
    parameters:
      - id: BMI
        component:
          id: bds_bmi
    ```

    Creates a new parameter (BMI) using the `bds_bmi` component. This typically derives BMI values from HEIGHT and WEIGHT parameters.

2.  Parameter operation with parameters

    ``` yml
    parameters:
      - id: BMI
        component:
          id: bds_bmi
          with:
            height: HEIGHT
            weight: WEIGHT
            by: [USUBJID, AVISITN]
    ```

    Creates BMI parameter with explicit specification of input parameters and grouping variables.

3.  Parameter operation with dependencies

    ``` yml
    parameters:
      - id: BMI
        component:
          id: bds_bmi
      - id: BMI_CATEGORY
        component:
          id: bds_groups
        depends:
          - parameters.BMI
        with:
          input: BMI
          output: BMIGRP
          bins: [0, 20, 25, 30, 100]
          labels: ["<20", "20-<25", "25-<30", "30+"]
    ```

    Creates a BMI category parameter that depends on the BMI parameter being derived first. Note the `parameters.` prefix in the dependency.

4.  Multiple parameters using the same component

    ``` yml
    parameters:
      - id: PROTEIN
        component:
          id: add_lab_parameter
          with:
            test_name: Protein
      - id: ALBUMIN
        component:
          id: add_lab_parameter
          with:
            test_name: Albumin
    ```

    Uses the same component to add multiple lab parameters with different test names.

# Row vs. parameter operations

The difference between row and parameter operations is entirely for the benefit of generating the define.xml document with mighty.toolbox. In mighty both operate the same way and follow the same rules for the ADaM specifications.
