---
title: "Get started"
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| include: false
#| eval: true

library(mighty)

temp_dir <- file.path(tempdir(), "mighty_example")
dir.create(temp_dir, showWarnings = FALSE, recursive = TRUE)
assign("temp_dir", temp_dir, envir = .GlobalEnv)


trial_yaml <- 'trial_id: "0001"
project_id: "0001"
complete_id: "0001-0001"
instance: "current"

keys:
  DM: [USUBJID, STUDYID]
  LB: [USUBJID, STUDYID, LBSEQ]
'
writeLines(trial_yaml, file.path(temp_dir, "example_01_trial.yaml"))
assign("trial_yaml", trial_yaml, envir = .GlobalEnv)


adsl_yaml <- 'id: ADSL
label: Subject Level Analysis Dataset
class: SUBJECT LEVEL ANALYSIS DATASET
structure: One record per subject
keys: [USUBJID, STUDYID]

population:
  base:
    - domain: DM
      depends:
        - AGE
      filter:
        - \'!is.na(AGE)\'
  global:
    - filter: NA
      depends:
        - NA

columns:
  - id: USUBJID

  - id: STUDYID

  - id: ARM

  - id: PLANNED_ARM
    method: ARM

  - id: AGE

  - id: AGE_GRP1
    component:
      id: "age_group_01.R"
'
# The path to the age_group.R component needs to be an absolute path when
# the vignette is rendered by `pkgdown`, but we don't want to show
# that ugly tmp path to the users
adsl_yaml_with_abspath <- gsub(
  '"age_group_01.R"',
  sprintf('"%s"', normalizePath(file.path(temp_dir, "age_group_01.R"), winslash = "/", mustWork = FALSE)),
  adsl_yaml
)
writeLines(adsl_yaml_with_abspath, file.path(temp_dir, "example_01_adsl.yaml"))
assign("adsl_yaml", adsl_yaml, envir = .GlobalEnv)


component_code <- "#' @title Age group 01
#' @description Grouping of age
#' @type derivation
#' @depends ADSL AGE
#' @outputs AGE_GRP1
#' @returns `ADSL`
#' @code
ADSL <- ADSL |>
  dplyr::mutate(
    AGE_GRP1 = cut(
      AGE,
      breaks = c(-Inf, 18, 65, 75, Inf),
      labels = c(\"< 18 years\", \"18 - 64 years\", \"65 - 74 years\", \">= 75 years\"),
      right = FALSE
    )
  )
"
writeLines(component_code, file.path(temp_dir, "age_group_01.R"))
assign("component_code", component_code, envir = .GlobalEnv)


main_script_content <- "# ADSL with 1 base domain
#   - 4 predecessors
#   - 1 renamed predecessor
#   - 1 derivation

# Load mighty
library(mighty)

# Generate ADaM program
results <- mighty::generate_adam_code(
  adam_specifications = \"example_01_adsl.yaml\",
  path_trial_metadata = \"example_01_trial.yaml\",
  standards_lib = \"mighty.standards\",
  path_trial = \".\"
)

# Print rendered ADaM program
results$programs$`1_ADSL` |> cat()
"
writeLines(main_script_content, file.path(temp_dir, "example_01_main.R"))
assign("main_script_content", main_script_content, envir = .GlobalEnv)
```

# Your first mighty ADaM program

Here is a minimal example that demonstrates how to specify and render a simple ADSL program with mighty to get you acquainted with the core concepts of the framework.

## Study directory guidelines

Mighty does not have any conditions on how your directory is structured. The only constraints are, there needs to be:

- A location to store study-specific custom components
- A [connector config file](https://novonordisk-opensource.github.io/connector/) that specifies where your trial data is located (SDTM, metadata, and ADaM). Mighty expects it to be located under `path_trial` with the name `_connector.yml`.
- A location where the compiled ADaM programs can be written out to

## Set up your trial metadata file

This YAML file contains information that spans across all ADaM domains.
This includes trial information and primary keys in different domains that are used for joining tables.

For this example, we will focus on a simple ADSL table and only provide keys related to that domain.

**example_01_trial.yaml**

```{r}
#| echo: false
#| eval: true
#| results: asis

cat("```yml\n")
cat(trial_yaml)
cat("```")
```

## Set up your ADaM specification

Provide a YAML file that specifies which columns to include in ADSL and how they are derived.

This ADSL table contains 6 columns: 4 simple columns from the base domain, 1 renamed column, and 1 derivation using a custom component.

**example_01_adsl.yaml**

```{r}
#| echo: false
#| eval: true
#| results: asis

cat("```yml\n")
cat(adsl_yaml)
cat("```")
```

## Define your custom component

The column `AGE_GRP1` in ADSL is derived by using a custom component that contains the logic to process the ADSL table by deriving the additional column `AGE_GRP1`.

**age_group_01.R**

```{r}
#| echo: false
#| eval: true
#| results: asis

cat("```r\n")
cat(component_code)
cat("```")
```

## Render the ADaM program

Call mighty to consolidate, analyze and render your ADSL program:

**example_01_main.R**

```{r}
#| echo: false
#| eval: true
#| results: asis

cat("```r\n")
cat(main_script_content)
cat("\n```")
```

Printing the rendered ADaM program yields a complete program for generating ADSL based on the specifications

```{r}
#| echo: false
#| eval: true
#| results: asis

# Execute the code generation
results <- mighty::generate_adam_code(
  adam_specifications = file.path(temp_dir, "example_01_adsl.yaml"),
  path_trial_metadata = file.path(temp_dir, "example_01_trial.yaml"),
  standards_lib = "mighty.standards",
  path_trial = temp_dir
)

# Display the generated program with R formatting
cat("```r\n")
cat(results$programs$`1_ADSL`)
cat("\n```")
```
