---
title: "Organizing ADaM programs"
vignette: >
  %\VignetteIndexEntry{Organizing ADaM programs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    embed-resources: true
    mermaid:
      theme: forest
---
# Overview

Explain topology analysis in broad terms

# Example 1

Example is found in Gewtting started.

```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    DM[("DM")]
    
    subgraph A_group [" "]
        direction LR
        A["Initialize ADSL"]
        A_output["USUBJID<br/>STUDYID<br/>ARM<br/>AGE"]
    end
    
    subgraph B_group [" "]
        direction LR
        B["age_group_01"]
        B_output["AGE_GRP1"]
    end
    
    subgraph C_group [" "]
        direction LR
        C["Copy"]
        C_output["PLANNED_ARM"]
    end
    
    DM --> A
    A -- AGE --> B
    A -- ARM --> C
    
    style DM fill:#2196F3,stroke:#1976D2,color:#fff
    style A_group fill:none,stroke:none,margin:0px,padding:2px
    style B_group fill:none,stroke:none,margin:0px,padding:2px
    style C_group fill:none,stroke:none,margin:0px,padding:2px
    style A fill:#4CAF50,stroke:#2E7D32,color:#fff
    style B fill:#4CAF50,stroke:#2E7D32,color:#fff
    style C fill:#4CAF50,stroke:#2E7D32,color:#fff
    style A_output fill:#FFC107,stroke:#F57C00,color:#000
    style B_output fill:#FFC107,stroke:#F57C00,color:#000
    style C_output fill:#FFC107,stroke:#F57C00,color:#000
```

# Example 2

```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    subgraph data_sources [" "]
        direction LR
        DM[("DM")]
        DM_VACCINE[("DM_VACCINE")]
    end
    
    subgraph A_group [" "]
        direction LR
        A["Initialize ADSL"]
        A_output["USUBJID<br/>STUDYID<br/>ARM<br/>AGE"]
    end
    
    subgraph B_group [" "]
        direction LR
        B["age_group_01"]
        B_output["AGE_GRP1"]
    end
    
    subgraph C_group [" "]
        direction LR
        C["Copy"]
        C_output["PLANNED_ARM"]
    end
    
    subgraph D_group [" "]
        direction LR
        D["age_category_01"]
        D_output["AGE_CAT1"]
    end
    
    data_sources --> A
    A -- AGE --> B
    A -- ARM --> C
    B -- AGE_GRP1 --> D
    
    style data_sources fill:#E3F2FD,stroke:#1976D2,stroke-width:2px
    style DM fill:#2196F3,stroke:#1976D2,color:#fff
    style DM_VACCINE fill:#2196F3,stroke:#1976D2,color:#fff
    style A_group fill:none,stroke:none,margin:0px,padding:2px
    style B_group fill:none,stroke:none,margin:0px,padding:2px
    style C_group fill:none,stroke:none,margin:0px,padding:2px
    style D_group fill:none,stroke:none,margin:0px,padding:2px
    style A fill:#4CAF50,stroke:#2E7D32,color:#fff
    style B fill:#4CAF50,stroke:#2E7D32,color:#fff
    style C fill:#4CAF50,stroke:#2E7D32,color:#fff
    style D fill:#4CAF50,stroke:#2E7D32,color:#fff
    style A_output fill:#FFC107,stroke:#F57C00,color:#000
    style B_output fill:#FFC107,stroke:#F57C00,color:#000
    style C_output fill:#FFC107,stroke:#F57C00,color:#000
    style D_output fill:#FFC107,stroke:#F57C00,color:#000
```

**example_02_trial.yaml**

```{yml, eval=FALSE}
trial_id: "0001"
project_id: "0001"
complete_id: "0001-0001"
instance: "current"

keys:
  DM: [USUBJID, STUDYID]
```

**example_02_adsl.yaml**

```{yml, eval=FALSE}
id: ADSL

keys: [USUBJID, STUDYID]

population:
  base:
    - domain: DM
      depends:
        - AGE
      filter:
        - '!is.na(AGE)'
    - domain: DM_VACCINE
      depends: NA
      filter: NA
  global:
    - filter: NA
      depends:
        - NA

columns:
  - id: USUBJID

  - id: STUDYID

  - id: ARM

  - id: PLANNED_ARM
    method: ARM

  - id: AGE

  - id: AGE_GRP1
    component:
      id: "components/age_group_01.R"

  - id: AGE_CAT1
    component:
      id: "components/age_category_01.R"
```

**age_category_01.R**

```{r, eval=FALSE}
#' @title Age category 01
#' @description Categorization of age group
#' @type derivation
#' @depends ADSL AGE_GRP1
#' @outputs AGE_CAT1
#' @returns `ADSL`
#' @code

ADSL <-  ADSL |> dplyr::mutate(
  AGE_CAT1 = dplyr::case_when(
    AGE_GRP1 %in% c("< 18 years", ">= 75 years") ~ "Outside expected range",
    TRUE                           ~ "Inside expected range"
  )
)

```

**example_02_main.R**

```{r, eval=FALSE}
## ADSL with 2 base domains
#   - 4 predecessors
#   - 1 renamed predecessor
#   - 2 derivation, with one depending on the other

# Load mighty
library(mighty)

# Generate ADaM program
results <- mighty::generate_adam_code(adam_specifications = "example_02_adsl.yaml",
  path_trial_metadata = file.path("example_02_trial.yaml"),
  standards_lib = "mighty.standards",
  path_trial = "."
)

# Print rendered ADaM program
results$programs$`1_ADSL` |> cat()
```

Printing the rendered ADaM program yields a complete program for generating ADSL based on the specification:s

```{r, eval=FALSE}
# ADSL-1-read_data -------------------------------------------------------------
cnt <- connector::connect(config = './_connector.yml')
  dm <-  cnt$sdtm$read_cnt('dm') |>
  dplyr::select(AGE, ARM, STUDYID, USUBJID)
  dm_vaccine <-  cnt$sdtm$read_cnt('dm_vaccine') |>
  dplyr::select(AGE, ARM, STUDYID, USUBJID)

# ADSL-init_domain -------------------------------------------------------------
dm <- dm |>
  dplyr::mutate(SRC_ = "dm")

dm_vaccine <- dm_vaccine |>
  dplyr::mutate(SRC_ = "dm_vaccine")

ADSL <-  rbind(dm,
dm_vaccine) |>
    dplyr::select(AGE,ARM,STUDYID,USUBJID,SRC_) |>
    admiral::convert_blanks_to_na()

# ADSL-filter_domain -----------------------------------------------------------
ADSL <-  ADSL |>
  dplyr::filter((SRC_ == 'dm' & !is.na(AGE)) | (SRC_ == 'dm_vaccine')) |>
  dplyr::select(-SRC_)

ADSL <-  ADSL |>
  dplyr::select(AGE, ARM, STUDYID, USUBJID)


# ADSL-AGE_GRP1 ----------------------------------------------------------------

  ADSL <-  ADSL |>
    dplyr::mutate(
      AGE_GRP1 = cut(
        AGE,
        breaks = c(-Inf, 18, 65, 75, Inf),
        labels = c("< 18 years", "18 - 64 years", "65 - 74 years", ">= 75 years"),
        right = FALSE
      )
    )
  

# ADSL-AGE_CAT1 ----------------------------------------------------------------

  ADSL <-  ADSL |> dplyr::mutate(
    AGE_CAT1 = dplyr::case_when(
      AGE_GRP1 %in% c("< 18 years", ">= 75 years") ~ "Outside expected range",
      TRUE                           ~ "Inside expected range"
    )
  )
  

# ADSL-PLANNED_ARM -------------------------------------------------------------
ADSL <- ADSL |> dplyr::mutate(PLANNED_ARM = ARM)

# ADSL-1-write_data ------------------------------------------------------------
# Sort rows by primary key
ADSL <- ADSL |> dplyr::arrange(USUBJID,STUDYID)

# Sort columns
ADSL <- ADSL |> dplyr::select(USUBJID,
STUDYID,
ARM,
PLANNED_ARM,
AGE,
AGE_GRP1,
AGE_CAT1)

# Save ADaM table
cnt$adam$write_cnt(ADSL, "ADSL.parquet", overwrite = TRUE)
```


# Example 3

```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    subgraph data_sources [" "]
        direction LR
        DM[("DM")]
        DM_VACCINE[("DM_VACCINE")]
    end
    
    subgraph A_group [" "]
        direction LR
        A["Initialize ADSL"]
        A_output["USUBJID<br/>STUDYID<br/>ARM<br/>AGE"]
    end
    
    subgraph B_group [" "]
        direction LR
        B["age_group_01"]
        B_output["AGE_GRP1"]
    end
    
    subgraph C_group [" "]
        direction LR
        C["Copy"]
        C_output["PLANNED_ARM"]
    end
    
    subgraph D_group [" "]
        direction LR
        D["age_category_01"]
        D_output["AGE_CAT1"]
    end
    
    data_sources --> A
    A -- AGE --> B
    A -- ARM --> C
    B -- AGE_GRP1 --> D
    
    style data_sources fill:#E3F2FD,stroke:#1976D2,stroke-width:2px
    style DM fill:#2196F3,stroke:#1976D2,color:#fff
    style DM_VACCINE fill:#2196F3,stroke:#1976D2,color:#fff
    style A_group fill:none,stroke:none,margin:0px,padding:2px
    style B_group fill:none,stroke:none,margin:0px,padding:2px
    style C_group fill:none,stroke:none,margin:0px,padding:2px
    style D_group fill:none,stroke:none,margin:0px,padding:2px
    style A fill:#4CAF50,stroke:#2E7D32,color:#fff
    style B fill:#4CAF50,stroke:#2E7D32,color:#fff
    style C fill:#4CAF50,stroke:#2E7D32,color:#fff
    style D fill:#4CAF50,stroke:#2E7D32,color:#fff
    style A_output fill:#FFC107,stroke:#F57C00,color:#000
    style B_output fill:#FFC107,stroke:#F57C00,color:#000
    style C_output fill:#FFC107,stroke:#F57C00,color:#000
    style D_output fill:#FFC107,stroke:#F57C00,color:#000
```

```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    LB[("LB")]
    
    subgraph E_group [" "]
        direction LR
        E["Initialize ADLB"]
        E_output["USUBJID<br/>STUDYID<br/>LBSEQ<br/>LBSTRESN"]
    end
    
    subgraph F_group [" "]
        direction LR
        F["Copy"]
        F_output["AVAL"]
    end
    
    LB --> E
    E -- LBSTRESN --> F
    
    style LB fill:#2196F3,stroke:#1976D2,color:#fff
    style E_group fill:none,stroke:none,margin:0px,padding:2px
    style F_group fill:none,stroke:none,margin:0px,padding:2px
    style E fill:#4CAF50,stroke:#2E7D32,color:#fff
    style F fill:#4CAF50,stroke:#2E7D32,color:#fff
    style E_output fill:#FFC107,stroke:#F57C00,color:#000
    style F_output fill:#FFC107,stroke:#F57C00,color:#000
```


```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    %% First flowchart - ADSL processing
    subgraph ADSL_flow ["**1_ADSL**"]
        subgraph data_sources [" "]
            direction LR
            DM[("DM")]
            DM_VACCINE[("DM_VACCINE")]
        end
        
        subgraph A_group [" "]
            direction LR
            A["Initialize ADSL"]
            A_output["USUBJID<br/>STUDYID<br/>ARM<br/>AGE"]
        end
        
        subgraph B_group [" "]
            direction LR
            B["age_group_01"]
            B_output["AGE_GRP1"]
        end
        
        subgraph C_group [" "]
            direction LR
            C["Copy"]
            C_output["PLANNED_ARM"]
        end
        
        subgraph D_group [" "]
            direction LR
            D["age_category_01"]
            D_output["AGE_CAT1"]
        end
        
        data_sources --> A
        A -- AGE --> B
        A -- ARM --> C
        B -- AGE_GRP1 --> D
    end
    
    %% Second flowchart - ADLB processing
    subgraph ADLB_flow ["**2_ADLB**"]
        LB[("LB")]
        
        subgraph E_group [" "]
            direction LR
            E["Initialize ADLB"]
            E_output["USUBJID<br/>STUDYID<br/>LBSEQ<br/>LBSTRESN"]
        end
        
        subgraph F_group [" "]
            direction LR
            F["Copy"]
            F_output["AVAL"]
        end
        
        LB --> E
        E -- LBSTRESN --> F
    end
    
    %% Third flowchart - AVAL_MIN processing
    subgraph ADSL_update ["**3_ADSL**"]
        subgraph G_group [" "]
            direction LR
            G["minimum_aval_01"]
            G_output["AVAL_MIN"]
        end
        
        A -- USUBJID --> G
        E -- USUBJID --> G
        F -- AVAL --> G
    end
    
    %% Styling
    style ADSL_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style ADLB_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style ADSL_update fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    
    style data_sources fill:#E3F2FD,stroke:#1976D2,stroke-width:2px
    style DM fill:#2196F3,stroke:#1976D2,color:#fff
    style DM_VACCINE fill:#2196F3,stroke:#1976D2,color:#fff
    style LB fill:#2196F3,stroke:#1976D2,color:#fff
    
    style A_group fill:none,stroke:none,margin:0px,padding:2px
    style B_group fill:none,stroke:none,margin:0px,padding:2px
    style C_group fill:none,stroke:none,margin:0px,padding:2px
    style D_group fill:none,stroke:none,margin:0px,padding:2px
    style E_group fill:none,stroke:none,margin:0px,padding:2px
    style F_group fill:none,stroke:none,margin:0px,padding:2px
    style G_group fill:none,stroke:none,margin:0px,padding:2px
    
    style A fill:#4CAF50,stroke:#2E7D32,color:#fff
    style B fill:#4CAF50,stroke:#2E7D32,color:#fff
    style C fill:#4CAF50,stroke:#2E7D32,color:#fff
    style D fill:#4CAF50,stroke:#2E7D32,color:#fff
    style E fill:#4CAF50,stroke:#2E7D32,color:#fff
    style F fill:#4CAF50,stroke:#2E7D32,color:#fff
    style G fill:#4CAF50,stroke:#2E7D32,color:#fff
    
    style A_output fill:#FFC107,stroke:#F57C00,color:#000
    style B_output fill:#FFC107,stroke:#F57C00,color:#000
    style C_output fill:#FFC107,stroke:#F57C00,color:#000
    style D_output fill:#FFC107,stroke:#F57C00,color:#000
    style E_output fill:#FFC107,stroke:#F57C00,color:#000
    style F_output fill:#FFC107,stroke:#F57C00,color:#000
    style G_output fill:#FFC107,stroke:#F57C00,color:#000
```

```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    %% First flowchart - ADSL processing
    subgraph ADSL_flow [" "]
        ADSL_title["<div  style='text-align: left; font-weight: bold;'>ADSL Processing</div>"]
        
        subgraph data_sources [" "]
            direction LR
            DM[("DM")]
            DM_VACCINE[("DM_VACCINE")]
        end
        
        subgraph A_group [" "]
            direction LR
            A["Initialize ADSL"]
            A_output["USUBJID<br/>STUDYID<br/>ARM<br/>AGE"]
        end
        
        subgraph B_group [" "]
            direction LR
            B["age_group_01"]
            B_output["AGE_GRP1"]
        end
        
        subgraph C_group [" "]
            direction LR
            C["Copy"]
            C_output["PLANNED_ARM"]
        end
        
        subgraph D_group [" "]
            direction LR
            D["age_category_01"]
            D_output["AGE_CAT1"]
        end
        
        ADSL_title ~~~ data_sources
        data_sources --> A
        A -- AGE --> B
        A -- ARM --> C
        B -- AGE_GRP1 --> D
    end
    
    %% Second flowchart - ADLB processing
    subgraph ADLB_flow [" "]
        ADLB_title["<div  style='text-align: left; font-weight: bold;'>ADLB Processing</div>"]
        
        LB[("LB")]
        
        subgraph E_group [" "]
            direction LR
            E["Initialize ADLB"]
            E_output["USUBJID<br/>STUDYID<br/>LBSEQ<br/>LBSTRESN"]
        end
        
        subgraph F_group [" "]
            direction LR
            F["Copy"]
            F_output["AVAL"]
        end
        
        ADLB_title ~~~ LB
        LB --> E
        E -- LBSTRESN --> F
    end
    
    %% Third flowchart - AVAL_MIN processing
    subgraph AVAL_MIN_flow [" "]
        AVAL_MIN_title["<div  style='text-align: left; font-weight: bold;'>AVAL_MIN Processing</div>"]
        
        subgraph G_group [" "]
            direction LR
            G["Calculate AVAL_MIN"]
            G_output["AVAL_MIN"]
        end
        
        AVAL_MIN_title ~~~ G_group
    end
    
    %% Cross-flow connections
    ADSL_flow --> AVAL_MIN_flow
    ADLB_flow --> AVAL_MIN_flow
    
    %% Styling
    style ADSL_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style ADLB_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style AVAL_MIN_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    
    style ADSL_title fill:none,stroke:none,color:#333
    style ADLB_title fill:none,stroke:none,color:#333
    style AVAL_MIN_title fill:none,stroke:none,color:#333
    
    style data_sources fill:#E3F2FD,stroke:#1976D2,stroke-width:2px
    style DM fill:#2196F3,stroke:#1976D2,color:#fff
    style DM_VACCINE fill:#2196F3,stroke:#1976D2,color:#fff
    style LB fill:#2196F3,stroke:#1976D2,color:#fff
    
    style A_group fill:none,stroke:none,margin:0px,padding:2px
    style B_group fill:none,stroke:none,margin:0px,padding:2px
    style C_group fill:none,stroke:none,margin:0px,padding:2px
    style D_group fill:none,stroke:none,margin:0px,padding:2px
    style E_group fill:none,stroke:none,margin:0px,padding:2px
    style F_group fill:none,stroke:none,margin:0px,padding:2px
    style G_group fill:none,stroke:none,margin:0px,padding:2px
    
    style A fill:#4CAF50,stroke:#2E7D32,color:#fff
    style B fill:#4CAF50,stroke:#2E7D32,color:#fff
    style C fill:#4CAF50,stroke:#2E7D32,color:#fff
    style D fill:#4CAF50,stroke:#2E7D32,color:#fff
    style E fill:#4CAF50,stroke:#2E7D32,color:#fff
    style F fill:#4CAF50,stroke:#2E7D32,color:#fff
    style G fill:#4CAF50,stroke:#2E7D32,color:#fff
    
    style A_output fill:#FFC107,stroke:#F57C00,color:#000
    style B_output fill:#FFC107,stroke:#F57C00,color:#000
    style C_output fill:#FFC107,stroke:#F57C00,color:#000
    style D_output fill:#FFC107,stroke:#F57C00,color:#000
    style E_output fill:#FFC107,stroke:#F57C00,color:#000
    style F_output fill:#FFC107,stroke:#F57C00,color:#000
    style G_output fill:#FFC107,stroke:#F57C00,color:#000
```

```{mermaid}
%%{init: {'flowchart': {'nodeSpacing': 5, 'rankSpacing': 50}, 'theme': 'base', 'themeVariables': {'edgeLabelBackground': '#ffffff', 'primaryTextColor': '#000000', 'secondaryTextColor': '#000000', 'tertiaryTextColor': '#000000'}}}%%
flowchart TB
    %% First flowchart - ADSL processing
    subgraph ADSL_flow ["ADSL Processing"]
        direction TB
        
        subgraph data_sources [" "]
            direction LR
            DM[("DM")]
            DM_VACCINE[("DM_VACCINE")]
        end
        
        subgraph A_group [" "]
            direction LR
            A["Initialize ADSL"]
            A_output["USUBJID<br/>STUDYID<br/>ARM<br/>AGE"]
        end
        
        subgraph B_group [" "]
            direction LR
            B["age_group_01"]
            B_output["AGE_GRP1"]
        end
        
        subgraph C_group [" "]
            direction LR
            C["Copy"]
            C_output["PLANNED_ARM"]
        end
        
        subgraph D_group [" "]
            direction LR
            D["age_category_01"]
            D_output["AGE_CAT1"]
        end
        
        data_sources --> A
        A -- AGE --> B
        A -- ARM --> C
        B -- AGE_GRP1 --> D
    end
    
    %% Second flowchart - ADLB processing
    subgraph ADLB_flow ["ADLB Processing"]
        direction TB
        
        LB[("LB")]
        
        subgraph E_group [" "]
            direction LR
            E["Initialize ADLB"]
            E_output["USUBJID<br/>STUDYID<br/>LBSEQ<br/>LBSTRESN"]
        end
        
        subgraph F_group [" "]
            direction LR
            F["Copy"]
            F_output["AVAL"]
        end
        
        LB --> E
        E -- LBSTRESN --> F
    end
    
    %% Third flowchart - AVAL_MIN processing
    subgraph AVAL_MIN_flow ["AVAL_MIN Processing"]
        direction TB
        
        subgraph G_group [" "]
            direction LR
            G["Calculate AVAL_MIN"]
            G_output["AVAL_MIN"]
        end
    end
    
    %% Cross-flow connections
    ADSL_flow --> AVAL_MIN_flow
    ADLB_flow --> AVAL_MIN_flow
    
    %% Styling
    style ADSL_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style ADLB_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style AVAL_MIN_flow fill:#f5f5f5,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    
    style data_sources fill:#E3F2FD,stroke:#1976D2,stroke-width:2px
    style DM fill:#2196F3,stroke:#1976D2,color:#fff
    style DM_VACCINE fill:#2196F3,stroke:#1976D2,color:#fff
    style LB fill:#2196F3,stroke:#1976D2,color:#fff
    
    style A_group fill:none,stroke:none,margin:0px,padding:2px
    style B_group fill:none,stroke:none,margin:0px,padding:2px
    style C_group fill:none,stroke:none,margin:0px,padding:2px
    style D_group fill:none,stroke:none,margin:0px,padding:2px
    style E_group fill:none,stroke:none,margin:0px,padding:2px
    style F_group fill:none,stroke:none,margin:0px,padding:2px
    style G_group fill:none,stroke:none,margin:0px,padding:2px
    
    style A fill:#4CAF50,stroke:#2E7D32,color:#fff
    style B fill:#4CAF50,stroke:#2E7D32,color:#fff
    style C fill:#4CAF50,stroke:#2E7D32,color:#fff
    style D fill:#4CAF50,stroke:#2E7D32,color:#fff
    style E fill:#4CAF50,stroke:#2E7D32,color:#fff
    style F fill:#4CAF50,stroke:#2E7D32,color:#fff
    style G fill:#4CAF50,stroke:#2E7D32,color:#fff
    
    style A_output fill:#FFC107,stroke:#F57C00,color:#000
    style B_output fill:#FFC107,stroke:#F57C00,color:#000
    style C_output fill:#FFC107,stroke:#F57C00,color:#000
    style D_output fill:#FFC107,stroke:#F57C00,color:#000
    style E_output fill:#FFC107,stroke:#F57C00,color:#000
    style F_output fill:#FFC107,stroke:#F57C00,color:#000
    style G_output fill:#FFC107,stroke:#F57C00,color:#000
```

**example_03_trial.yaml**

```{yml, eval=FALSE}
trial_id: "0001"
project_id: "0001"
complete_id: "0001-0001"
instance: "current"

keys:
  DM: [USUBJID, STUDYID]
  DM_VACCINE: [USUBJID, STUDYID]
  LB: [USUBJID, STUDYID, LBSEQ]
```

**example_03_adsl.yaml**

```{yml, eval=FALSE}
id: ADSL

keys: [USUBJID, STUDYID]

population:
  base:
    - domain: DM
      depends:
        - AGE
      filter:
        - '!is.na(AGE)'
    - domain: DM_VACCINE
      depends: NA
      filter: NA
  global:
    - filter: NA
      depends:
        - NA

columns:
  - id: USUBJID

  - id: STUDYID

  - id: ARM

  - id: PLANNED_ARM
    method: ARM

  - id: AGE

  - id: AGE_GRP1
    component:
      id: "components/age_group_01.R"

  - id: AGE_CAT1
    component:
      id: "components/age_category_01.R"

  - id: MIN_AVAL
    component:
      id: "components/minimum_aval_01.R"
```

**example_03_adlb.yaml**

```{yml, eval=FALSE}
id: ADLB

keys: [USUBJID, STUDYID, LBSEQ]

population:
  base:
    - domain: LB
      depends:
        - LBSTRESN
      filter:
        - '!is.na(LBSTRESN)'
  global:
    - filter: NA
      depends:
        - NA

columns:
  - id: USUBJID

  - id: STUDYID

  - id: LBSEQ

  - id: LBSTRESN

  - id: AVAL
    method: LBSTRESN
```


**minimum_aval_01.R**

```{r, eval=FALSE}
#' @title Minimum AVAL per subject
#' @description A description
#' @type derivation
#' @depends ADSL USUBJID
#' @depends ADLB USUBJID
#' @depends ADLB AVAL
#' @outputs MIN_AVAL
#' @returns `ADSL`
#' @code

# Extract the minimum AVAL value for each SUBJID
ADSL <- ADSL |>
  dplyr::left_join(
    ADLB |> dplyr::select(USUBJID, AVAL) |>
      dplyr::group_by(USUBJID) |>
      dplyr::summarise(MIN_AVAL = min(AVAL)) |>
      dplyr::ungroup()
    ,
    by = "USUBJID"
  )
```

**example_03_main.R**

```{r, eval=FALSE}
# ADSL with 2 base domains
#   - 4 predecessors
#   - 1 renamed predecessor
#   - 2 derivation, with one depending on the other
#   - 1 derivation depending on ADLB
# ADLB with 1 base domain
#   - 4 predecessors 
#   - 1 renamed predecessor

# Load mighty
library(mighty)

# Generate ADaM program
results <- mighty::generate_adam_code(adam_specifications = c("example_03_adsl.yaml",
                                                       "example_03_adlb.yaml"),
                                      path_trial_metadata = file.path("example_03_trial.yaml"),
                                      standards_lib = "mighty.standards",
                                      path_trial = "."
)


# Print names of rendered ADaM programs
results$programs |> names()

# Print rendered ADaM programs
results$programs$`1_ADSL` |> cat()
results$programs$`2_ADLB` |> cat()
results$programs$`3_ADSL` |> cat()
```

**1_ADSL**
```{r, eval=FALSE}
# ADSL-1-read_data -------------------------------------------------------------
cnt <- connector::connect(config = './_connector.yml')
  dm <-  cnt$sdtm$read_cnt('dm') |>
  dplyr::select(AGE, ARM, STUDYID, USUBJID)
  dm_vaccine <-  cnt$sdtm$read_cnt('dm_vaccine') |>
  dplyr::select(AGE, ARM, STUDYID, USUBJID)

# ADSL-init_domain -------------------------------------------------------------
dm <- dm |>
  dplyr::mutate(SRC_ = "dm")

dm_vaccine <- dm_vaccine |>
  dplyr::mutate(SRC_ = "dm_vaccine")

ADSL <-  rbind(dm,
dm_vaccine) |>
    dplyr::select(AGE,ARM,STUDYID,USUBJID,SRC_) |>
    admiral::convert_blanks_to_na()

# ADSL-filter_domain -----------------------------------------------------------
ADSL <-  ADSL |>
  dplyr::filter((SRC_ == 'dm' & !is.na(AGE)) | (SRC_ == 'dm_vaccine')) |>
  dplyr::select(-SRC_)

ADSL <-  ADSL |>
  dplyr::select(AGE, ARM, STUDYID, USUBJID)


# ADSL-AGE_GRP1 ----------------------------------------------------------------

  ADSL <-  ADSL |>
    dplyr::mutate(
      AGE_GRP1 = cut(
        AGE,
        breaks = c(-Inf, 18, 65, 75, Inf),
        labels = c("< 18 years", "18 - 64 years", "65 - 74 years", ">= 75 years"),
        right = FALSE
      )
    )
  

# ADSL-AGE_CAT1 ----------------------------------------------------------------

  ADSL <-  ADSL |> dplyr::mutate(
    AGE_CAT1 = dplyr::case_when(
      AGE_GRP1 %in% c("< 18 years", ">= 75 years") ~ "Outside expected range",
      TRUE                           ~ "Inside expected range"
    )
  )
  

# ADSL-PLANNED_ARM -------------------------------------------------------------
ADSL <- ADSL |> dplyr::mutate(PLANNED_ARM = ARM)

# ADSL-1-write_data ------------------------------------------------------------
# Sort rows by primary key
ADSL <- ADSL |> dplyr::arrange(USUBJID,STUDYID)

# Save ADaM table
cnt$adam$write_cnt(ADSL, "ADSL.parquet", overwrite = TRUE)
```

**2_ADLB**
```{r, eval=FALSE}
# ADLB-2-read_data -------------------------------------------------------------
cnt <- connector::connect(config = './_connector.yml')
  lb <-  cnt$sdtm$read_cnt('lb') |>
  dplyr::select(LBSEQ, LBSTRESN, STUDYID, USUBJID)

# ADLB-init_domain -------------------------------------------------------------
ADLB <-  lb |>
    dplyr::select(LBSEQ,LBSTRESN,STUDYID,USUBJID) |>
    admiral::convert_blanks_to_na()

# ADLB-filter_domain -----------------------------------------------------------
ADLB <-  ADLB |>
  dplyr::filter(!is.na(LBSTRESN))

ADLB <-  ADLB |>
  dplyr::select(LBSEQ, LBSTRESN, STUDYID, USUBJID)


# ADLB-AVAL --------------------------------------------------------------------
ADLB <- ADLB |> dplyr::mutate(AVAL = LBSTRESN)

# ADLB-2-write_data ------------------------------------------------------------
# Sort rows by primary key
ADLB <- ADLB |> dplyr::arrange(USUBJID,STUDYID,LBSEQ)

# Sort columns
ADLB <- ADLB |> dplyr::select(USUBJID,
STUDYID,
LBSEQ,
LBSTRESN,
AVAL)

# Save ADaM table
cnt$adam$write_cnt(ADLB, "ADLB.parquet", overwrite = TRUE)
```

**3_ADSL**
```{r, eval=FALSE}
# ADSL-3-read_data -------------------------------------------------------------
cnt <- connector::connect(config = './_connector.yml')
  ADSL <-  cnt$adam$read_cnt('ADSL')
  ADLB <-  cnt$adam$read_cnt('ADLB') |>
  dplyr::select(AVAL, USUBJID)

# ADSL-MIN_AVAL ----------------------------------------------------------------

  # Extract the minimum AVAL value for each SUBJID
  ADSL <- ADSL |>
    dplyr::left_join(
      ADLB |> dplyr::select(USUBJID, AVAL) |>
        dplyr::group_by(USUBJID) |>
        dplyr::summarise(MIN_AVAL = min(AVAL)) |>
        dplyr::ungroup()
      ,
      by = "USUBJID"
    )
  

# ADSL-3-write_data ------------------------------------------------------------
# Sort rows by primary key
ADSL <- ADSL |> dplyr::arrange(USUBJID,STUDYID)

# Sort columns
ADSL <- ADSL |> dplyr::select(USUBJID,
STUDYID,
ARM,
PLANNED_ARM,
AGE,
AGE_GRP1,
AGE_CAT1,
MIN_AVAL)

# Save ADaM table
cnt$adam$write_cnt(ADSL, "ADSL.parquet", overwrite = TRUE)
```
