---
title: "Special components"
vignette: >
  %\VignetteIndexEntry{Special components}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Special components are Mustache templates prefixed with `_` that mighty inserts
automatically into every generated ADaM program. Users never reference them in
YAML specs. This vignette documents the parameter contract for each special
component so that template authors can understand (and modify) the templates.

```{r}
#| label: setup
#| include: false
library(mighty.component)

component_path <- function(name) {
  p <- system.file(paste0("components/", name), package = "mighty")
  if (nzchar(p)) p else file.path("..", "inst", "components", name)
}

.html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x
}

component_detail <- function(name, params = NULL, renders = NULL) {
  template_html <- .html_escape(paste(readLines(component_path(name)), collapse = "\n"))

  if (is.null(renders)) {
    render_html <- paste0(
      "<p><strong>Rendered output</strong></p>\n",
      "<pre><code class=\"language-r\">",
      get_rendered_component(component_path(name), params)$code |>
        paste(collapse = "\n") |>
        .html_escape(),
      "</code></pre>\n"
    )
  } else {
    render_html <- paste(sapply(renders, function(r) {
      paste0(
        "<p><strong>", r$label, "</strong></p>\n",
        "<pre><code class=\"language-r\">",
        get_rendered_component(component_path(name), r$params)$code |>
          paste(collapse = "\n") |>
          .html_escape(),
        "</code></pre>\n"
      )
    }), collapse = "")
  }

  knitr::asis_output(paste0(
    "<details><summary>Standard template and rendered output</summary>\n",
    "<p><strong>Template</strong></p>\n",
    "<pre><code>", template_html, "</code></pre>\n",
    render_html,
    "</details>\n"
  ))
}
```

# Three stages of code generation

Mighty produces ADaM programs through three distinct stages, each with its own
environment. Knowing which stage you're looking at tells you what values are
available and what the parameters mean.

1. **Mighty pipeline** — mighty reads YAML specs, resolves dependencies, and
   decides which components to use and in what order. This is where the
   parameter lists described below get built. The environment is mighty's own R
   session with full access to mighty internals.

2. **Template rendering** — mighty.component renders each Mustache template by
   substituting parameters into `{{placeholders}}`. The template sees only the
   named list that mighty passes to it.

3. **ADaM program execution** — the rendered R code runs later, in the user's
   own R session, as a standalone script. It has access to the `connector`
   package, `dplyr`, and whatever the user's project loads — but not to mighty
   or mighty.component.

This vignette documents the named lists that mighty passes at the Template
rendering stage.

# Template syntax

The parameter examples in later sections show values that get substituted into
Mustache placeholders. Here's how interpolation works in these templates.

Mustache has two interpolation forms:

| Syntax | Behavior |
|--------|-----------|
| `{{{variable}}}` | Inserts the value with HTML escaping — `"` becomes `&quot;`, `<` becomes `&lt;`, etc. |
| `{{{{variable}}}}` | Inserts the value verbatim — no escaping |

All variable interpolations in these templates use triple-mustache `{{{{variable}}}}`, in both `@code` sections and documentation headers (`@title`, `@description`, `@param`, `@depends`, `@outputs`). This avoids HTML escaping in generated code and defers any format-specific rendering to mighty.component.

The only exception is **section tags** (`{{{#condition}}}`, `{{{/condition}}}`, `{{{^condition}}}`), which are always double-mustache — this is a Mustache language requirement.

# The `keep_vars` pipeline

Before diving into each template, it helps to understand one parameter that
appears in four of them with different contents each time.

`keep_vars` appears in the four structural templates (`mighty_read_data`,
`mighty_init_domain`, `mighty_filter_domain`, `mighty_write_data`) but not in the column-level
templates (`mighty_col_mutate`, `mighty_col_rename`, `mighty_col_echo`). It carries a different
set of columns at each stage:

| Template | `keep_vars` contains | Purpose |
|---|---|---|
| `mighty_read_data` | All columns needed by any downstream step, per source domain | Limits memory footprint on read |
| `mighty_init_domain` | Same list as `mighty_read_data` | Selects columns after row binding |
| `mighty_filter_domain` | Columns surviving after filtering, minus `SRC_` and filter-only columns | Drops columns no longer needed post-filter |
| `mighty_write_data` | Final output columns in YAML spec order | Controls what is persisted and in what order |

`keep_vars` in `mighty_read_data` is the origin: every column referenced in later
templates must have been loaded there first. If a column appears in
`mighty_init_domain`, `mighty_filter_domain`, or `mighty_write_data` but was not included in
`mighty_read_data`, the generated program will fail at execution time with a column
not found error. Mighty computes all four lists from the same YAML specification
so this is guaranteed automatically — but it matters when modifying templates or
building custom pipelines.

# mighty_read_data

Rendered at the top of every ADaM program. Opens a connector and reads each
source dataset, selecting only the columns needed downstream.

## Example

Given this YAML specification:

```yaml
id: ADSL
keys:
  - USUBJID
  - STUDYID
population:
  base:
    - domain: DM
      depends: [NA]
      filter: NA
    - domain: DM_VACCINE
      depends: [NA]
      filter: NA
columns:
  - id: USUBJID
  - id: STUDYID
  - id: ARM
```

Mighty builds the following parameter list and passes it to the template:

```r
list(
  connector_path_expr = '"_connector.yml"',
  domains = list(
    list(
      is_current_domain = FALSE,
      domain_name = "DM",
      data_type = "sdtm",
      keep_vars = "ARM, STUDYID, USUBJID"
    ),
    list(
      is_current_domain = FALSE,
      domain_name = "DM_VACCINE",
      data_type = "sdtm",
      keep_vars = "ARM, STUDYID, USUBJID"
    )
  )
)
```

The standard template and a rendered version using the parameter list above can be expanded below.

```{r}
#| echo: false
component_detail("mighty_read_data", list(
  connector_path_expr = '"_connector.yml"',
  domains = list(
    list(is_current_domain = FALSE, domain_name = "DM", data_type = "sdtm", keep_vars = "ARM, STUDYID, USUBJID"),
    list(is_current_domain = FALSE, domain_name = "DM_VACCINE", data_type = "sdtm", keep_vars = "ARM, STUDYID, USUBJID")
  )
))
```

## Parameters

### `connector_path_expr`

A string containing an R expression that resolves to the connector config file
path at runtime. Two forms are possible:

| Config value | `connector_path_expr` |
|---|---|
| `_connector.yml` | `'"_connector.yml"'` (quoted literal) |
| `!expr here::here("_connector.yml")` | `'here::here("_connector.yml")'` (bare expression) |

The `!expr` prefix in the mighty config strips the prefix and passes the
remaining expression through verbatim. Plain paths are wrapped in double quotes
so they evaluate as character literals in the generated R code.

### `domains`

A list where each element represents one source dataset to read. Each element
is a named list with the following fields:

| Field | Type | Meaning |
|-------|------|---------|
| `domain_name` | character | Source dataset name, uppercase (e.g. `"DM"`, `"ADSL"`). |
| `data_type` | character | Which connector data source to use: `"sdtm"`, `"adam"`, or `"metadata"`. |
| `keep_vars` | character | Comma-separated column names to select, sorted alphabetically and uppercased. Only used when `is_current_domain` is `FALSE`. Column order in the final dataset is controlled by `keep_vars` in `mighty_write_data`, not by this field. |
| `is_current_domain` | logical | `TRUE` when the domain being read is the same one being built. When `TRUE`, `keep_vars` should be ignored. |

#### How `data_type` is determined

Domain names are classified using these rules:

| Pattern | `data_type` |
|---------|-------------|
| Two-character names (`DM`, `LB`, `VS`, ...), `RELREC`, `DM_*`, `SUPP*` | `"sdtm"` |
| Names starting with `AD` (`ADSL`, `ADLB`, ...) | `"adam"` |
| Names starting with `MD` | `"metadata"` |

#### When `is_current_domain` is `TRUE`

This happens in multi-program domains where a later program reads the output of
an earlier program for the same domain (e.g. an ADSL "update" program). When
`is_current_domain` is `TRUE`, `keep_vars` should be ignored.

# mighty_init_domain

Rendered once per program, after `mighty_read_data`. Combines one or more source
datasets by row binding and selects the columns needed for the domain.

## Example

Using the same ADSL specification from `mighty_read_data`:

```r
list(
  self = "ADSL",
  keep_vars = "ARM, STUDYID, USUBJID",
  source_domain_rbind = "rbind(DM,\nDM_VACCINE)",
  src_mutations = list()
)
```

`src_mutations` is empty here because both source domains have `filter: NA`.
When domain-specific filters are defined, mighty populates it so that a
temporary `SRC_` column can be added before row binding (see
[ADaM specification — filter execution order](adam_specification.html#filter-execution-order)):

```r
src_mutations = list(
  list(domain = "DM"),
  list(domain = "DM_VACCINE")
)
```

The standard template and a rendered version using the parameter list above can be expanded below.

```{r}
#| echo: false
component_detail("mighty_init_domain", list(
  self = "ADSL",
  keep_vars = "ARM, STUDYID, USUBJID",
  source_domain_rbind = "rbind(DM,\nDM_VACCINE)",
  src_mutations = list()
))
```

## Parameters

### `self`

The name of the ADaM domain being built (e.g. `"ADSL"`, `"ADLB"`). Used as
the variable name on the left-hand side of the assignment.

### `keep_vars`

Comma-separated column names to retain after row binding. Pre-formatted in R
before being passed to the template.

### `source_domain_rbind`

A pre-formatted R expression that combines source datasets. A single source
domain is passed as its name directly; multiple domains are wrapped in an
`rbind()` call with line breaks for readability.

### `src_mutations`

A list of per-domain mutation specifications used to add a `SRC_` tracking
column before row binding. Each element has a single field:

| Field | Type | Meaning |
|-------|------|---------|
| `domain` | character | Name of the source domain to tag with `SRC_`. |

`src_mutations` is empty when `SRC_` is not in `keep_vars`.

`SRC_` is an internal column that mighty adds and removes automatically to
support domain-specific row filters. See the
[ADaM specification vignette](adam_specification.html#filter-execution-order) for a
full explanation of how filtering works across source domains.

# mighty_filter_domain

Rendered after `mighty_init_domain`. Applies joins, domain-specific filters, global
filters, and column selection to the initialized dataset.

## Example

Given this YAML specification:

```yaml
id: ADLB
keys:
  - USUBJID
  - STUDYID
population:
  base:
    - domain: LB
      depends: [NA]
      filter: NA
    - domain: XL
      depends: [LBCAT]
      filter: "LBCAT == 'CHEMISTRY'"
  global:
    - filter: "!is.na(SEX)"
      depends: [ADSL.SEX]
columns:
  - id: USUBJID
  - id: STUDYID
  - id: PARAMCD
  - id: LBCAT
  - id: SEX
    method: ADSL.SEX
```

Mighty builds the following parameter list and passes it to the template:

```r
list(
  self = "ADLB",
  joins = list(
    list(
      table = "ADSL",
      select_expr = "SEX, STUDYID, USUBJID",
      keys = '"STUDYID", "USUBJID"'
    )
  ),
  domain_filter = "(SRC_ == 'LB') | (SRC_ == 'XL' & LBCAT == 'CHEMISTRY')",
  global_filter = "!is.na(SEX)",
  keep_vars = "LBCAT, PARAMCD, STUDYID, USUBJID"
)
```

::: {.callout-note}
`SEX` appears in the global filter (`!is.na(SEX)`) and is declared in the YAML
spec, but it is absent from `keep_vars`. This is intentional: `SEX` originates
from `ADSL` (via `method: ADSL.SEX`) and is not present in the LB or XL source
data. It is joined in temporarily to evaluate the filter, then dropped. A
subsequent `mighty_col_echo` step re-adds `SEX` to ADLB permanently via a left join
with ADSL.
:::

The standard template and a rendered version using the parameter list above can be expanded below.

```{r}
#| echo: false
component_detail("mighty_filter_domain", list(
  self = "ADLB",
  joins = list(
    list(table = "ADSL", select_expr = "SEX, STUDYID, USUBJID", keys = '"STUDYID", "USUBJID"')
  ),
  domain_filter = "(SRC_ == 'LB') | (SRC_ == 'XL' & LBCAT == 'CHEMISTRY')",
  global_filter = "!is.na(SEX)",
  keep_vars = "LBCAT, PARAMCD, STUDYID, USUBJID"
))
```

## Parameters

### `self`

The name of the ADaM domain being filtered.

### `joins`

A list of join specifications. Each element is a named list:

| Field | Type | Meaning |
|-------|------|---------|
| `table` | character | Name of the dataset to join with. |
| `select_expr` | character | Comma-separated column names to select from the join dataset. |
| `keys` | character | Comma-separated quoted column names used as join keys. |

`joins` is empty when no filter depends on an external dataset.

### `domain_filter`

Pre-formatted R expression filtering on `SRC_` (e.g.
`"(SRC_ == 'LB') | (SRC_ == 'XL' & LBCAT == 'CHEMISTRY')"`). `NULL` when no
domain-specific filter is defined — whisker treats `NULL` as falsy so the filter
block is suppressed.

### `global_filter`

Raw filter expression string from the YAML spec (e.g. `"!is.na(SEX)"`). `NULL`
when no global filter is defined — the filter block is suppressed.

### `keep_vars`

A comma-separated string of column names to retain. `NULL` when no selection
is needed — whisker treats `NULL` as falsy so the `dplyr::select()` call is
suppressed. `SRC_` is excluded because it is only used during filtering.

# mighty_col_mutate

Rendered for each `col_copy` action — variables copied verbatim from one column
to another, typically with an ADaM-standard rename (e.g. `LBSTRESN` → `AVAL`).

## Example

Consider the following YAML specification snippet. `AVAL` is defined with
`method: LBSTRESN`, and `LBSTRESN` is also listed as its own column. Mighty
copies the values rather than renaming because the source column is retained
in the output:

```yaml
columns:
  - id: LBSTRESN
  - id: AVAL
    method: LBSTRESN
```

Mighty builds the following parameter list and passes it to the template:

```r
list(
  self = "ADLB",
  rename_var = "AVAL",
  source_var = "LBSTRESN"
)
```

The standard template and a rendered version using the parameter list above can be expanded below.

```{r}
#| echo: false
component_detail("mighty_col_mutate",
  list(self = "ADLB", rename_var = "AVAL", source_var = "LBSTRESN")
)
```

## Parameters

### `self`

The name of the dataset being modified.

### `rename_var`

The name of the new column to create.

### `source_var`

The name of the existing column whose values are copied.

# mighty_col_rename

Rendered for each `col_rename` action — renames an existing column in-place without copying values. Unlike `mighty_col_mutate`, which creates a new column via `dplyr::mutate()`, this template uses `dplyr::rename()` and leaves no extra column behind.

Both templates share the same parameter list and are served by the same internal function (`params_mutate_code`). They differ only in the R verb the template emits.

## Example

Consider the following YAML specification snippet. `SRCSEQ` is defined with
`method: LBSEQ`, and `LBSEQ` is **not** listed as its own column. Because
the source column has no independent entry, mighty renames it rather than
copying it:

```yaml
columns:
  - id: SRCSEQ
    method: LBSEQ
```

Mighty builds the following parameter list and passes it to the template:

```r
list(
  self = "ADLB",
  rename_var = "SRCSEQ",
  source_var = "LBSEQ"
)
```

The standard template and a rendered version using the parameter list above can be expanded below.

```{r}
#| echo: false
component_detail("mighty_col_rename",
  list(self = "ADLB", rename_var = "SRCSEQ", source_var = "LBSEQ")
)
```

## Parameters

### `self`

The name of the dataset being modified.

### `rename_var`

The new column name after renaming.

### `source_var`

The existing column to rename.

# mighty_col_echo

Rendered for each `col_echo` action — variables pulled in from a different
domain via a left join.

## Example

Consider the following YAML specification snippet. The dot notation `ADSL.SEX`
in `method:` tells mighty the column originates from another domain, which
triggers a left join rather than a direct copy:

```yaml
columns:
  - id: SEX
    method: ADSL.SEX
```

Mighty builds the following parameter list and passes it to the template:

```r
list(
  self = "ADLB",
  join_dataset = "ADSL",
  select_expr = "STUDYID, USUBJID, SEX",
  by_vars = '"STUDYID", "USUBJID"',
  needs_rename = FALSE,
  output_var = "SEX",
  var_to_add = "SEX"
)
```

This is the step that permanently adds `SEX` to ADLB — the same column that
`mighty_filter_domain` joined in temporarily to evaluate the global filter.

The standard template and a rendered version using the parameter list above can be expanded below.

```{r}
#| echo: false
component_detail("mighty_col_echo", list(
  self = "ADLB",
  join_dataset = "ADSL",
  select_expr = "STUDYID, USUBJID, SEX",
  by_vars = '"STUDYID", "USUBJID"',
  needs_rename = FALSE,
  output_var = "SEX",
  var_to_add = "SEX"
))
```

## Parameters

### `self`

The name of the primary dataset being modified.

### `join_dataset`

The name of the dataset to join with (right-hand side of the join).

### `select_expr`

Comma-separated column names to select from `join_dataset` before joining.
Includes both the join keys and the variable being added.

### `by_vars`

Comma-separated quoted column names used as join keys, ready for use in
`dplyr::left_join(by = c(...))`.

### `needs_rename`

`TRUE` when the variable name in the source domain differs from the desired
output name. When `TRUE`, a `dplyr::rename()` call is appended.

### `output_var` / `var_to_add`

`var_to_add` is the column name as it exists in `join_dataset`. `output_var` is
the desired column name in `self`. When `needs_rename` is `FALSE` they are
equal.

# mighty_write_data

Rendered at the end of every ADaM program. Sorts rows, selects columns in
specification order, and persists the dataset via the connector.

## Example

For the ADSL specification from the `mighty_read_data` example, mighty builds:

```r
list(
  self = "ADSL",
  file_ext = "parquet",
  row_order_vars = "USUBJID,\nSTUDYID",
  keep_vars = "USUBJID,\nSTUDYID,\nARM"
)
```

`keep_vars` is always pre-formatted as a `,\n`-separated string.

::: {.callout-note}
When mighty is run with source data available for validation, columns that
cannot be derived due to missing source data are prefixed with `# ` in the
formatted string, commenting them out of the `dplyr::select()` call. This
allows the generated program to run and produce a partial dataset rather than
failing outright. This behaviour is part of mighty's missing data handling and
does not affect the template itself.
:::


```r
list(
  self = "ADLB",
  file_ext = "parquet",
  row_order_vars = "USUBJID,\nSTUDYID",
  keep_vars = "USUBJID,\nSTUDYID,\nPARAMCD,\nAVAL"
)
```

The template and two rendered versions — one with a short column list and one with a longer list — can be expanded below.

```{r}
#| echo: false
component_detail("mighty_write_data",
  renders = list(
    list(
      label = "Rendered output — short column list",
      params = list(
        self = "ADSL",
        file_ext = "parquet",
        row_order_vars = "USUBJID,\nSTUDYID",
        keep_vars = "USUBJID,\nSTUDYID,\nARM"
      )
    ),
    list(
      label = "Rendered output — long column list",
      params = list(
        self = "ADLB",
        file_ext = "parquet",
        row_order_vars = "USUBJID,\nSTUDYID",
        keep_vars = "USUBJID,\nSTUDYID,\nPARAMCD,\nAVAL"
      )
    )
  )
)
```

## Parameters

### `self`

The name of the domain being built (e.g. `"ADSL"`, `"ADLB"`).

### `file_ext`

File extension for the persisted dataset. Defaults to `"parquet"`. Will be
configurable via `_mighty.yml` (global default) and per-domain YAML
(`table_metadata`) in a future release.

### `row_order_vars`

The domain's primary keys, formatted as a `,\n`-separated string. `NULL` when
the domain has no primary keys — whisker treats `NULL` as falsy so the sort
block is suppressed.

### `keep_vars`

The output columns in specification order. `NULL` for intermediate programs in
multi-program domains — the template uses `keep_vars` directly as the section
condition, so a `NULL` value suppresses the `dplyr::select()` call. Always
passed as a `,\n`-separated string. See the callout above regarding columns that
may be prefixed with `# `.
